---
title: "**Disease Progression Model in Diabetes**"
author: "Umesh Singh"
date: "2025-12-27"
# output: html_document
always_allow_html: true
output:
  html_document:
    df_print: paged
    theme: default
    toc: false
    code_folding: hide
    # css: style.css
    includes:
      # in_header: "/ARC_Logo.png"
  classoption: landscape
    # mainfont: Arial
  pdf_document: 
    keep_tex: true
    latex_engine: xelatex
  # word_document: default
  # github_document: default
  fig_caption: yes
  keep_md: yes
---

```{r setup, include=FALSE, warning= FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<div style="max-width: 960px; height: 420px; overflow: hidden; margin: 0 auto 0 auto; background: transparent;">
  <video autoplay loop muted playsinline width="100%" style="width: 100%; height: 100%; object-fit: cover; object-position: center; transform: scale(1.0, 1.0); transform-origin: center center; border: 0; outline: none; box-shadow: none; background: transparent;">
    <source src="dpm.mp4" type="video/mp4">
  </video>
</div>

<div style="margin-top: -1.5rem;"></div>

## Disease Progression Model{.tabset .tabset-fade .tabset-pills}
### Project

Aims of this project was **to develop a disease progression model for diabetes**.
In this project, the different states are defined according to the glycosylated hemoglobin levels:
- **State 1: Normoglycemia (HbA1c<5.7)** <br>
- **State 2: Pre-Diabetes (HbA1c 5.7-6.4)** <br>
- **States 3: Diabetes (HbA1c>6.4)** <br>

The attending practitioner monitor the status of subjects with diabetes periodically over time. The patient comes to the clinic a random number of times and at random times. Relevant covariate information is recorded at every visit. During a clinic visit, the glycosylated hemoglobin (HbA1c) level is assessed. The status of the subject is determined by the HbA1c level.  <br>
**Research Question**: <br>
- Given a specific state of the patient at the current visit, what will be her/his state at the next visit (months, years) and  <br>
- how will that state be influenced by her/his gender, race and Risk score.

### Cohort Building{.tabset}
#### Data {.tabset}
```{r 1, warning= FALSE, message=FALSE}
# detach("package:digest", unload = TRUE)
# install.packages("digest")
# packageVersion("digest")
library(digest)
library(readxl)
library(dplyr)
library(msm) # multistate modeling
library(plyr) # extract info from data
# options(repos="https://CRAN.R-project.org")
# install.packages("openxlsx")
# library(openxlsx)
remove(dpm)
dpm<-read_excel("E:\\ARC\\Projects\\Dis Prog Model\\dpm\\DPM1.xlsx")
dpm$datevisit <- as.Date(dpm$datevisit, format = "%m/%d/%Y" )
dpm$DOB <- as.Date(dpm$DOB, format = "%m/%d/%Y" )
dpm$X <- as.Date('2025-12-27')
dpm$LastVisit=(dpm$X-dpm$datevisit)

# dpm$ID_hash <- sapply(dpm$ID, function(x) digest(x, algo = "sha256", key = salt))
salt <- Sys.getenv("HASH_SALT")
dpm$ID_hashed <- sapply(dpm$ID,  function(x) digest(paste0(x, salt), "sha256"))

dpm$ID_hash <- substr(dpm$ID_hashed,
                      nchar(dpm$ID_hashed) - 5,
                      nchar(dpm$ID_hashed))

dpm %>% select(-c(ID,DOB,ID_hashed)) %>% head()
```
The project is based on data from 242 unique patients with a total of 726 visits. Relevant data were accessed from the EHR of the high-risk patients for diabetic mellitus who are enrolld at the Center. Deidentified data was used in the following analysis.

- Describing the Dataset
  +  ID_hash = Patient ID (hashed)
  +  age_years = age of the patient in years exact to the day of visit 
  +  gender1 = gender (0 = female; 1= male )
  +  status = state of diabetes (1= no diabetes; 2=prediabetes; 3=diabetes)
  +  cumint = cumulative duration in months from the first visit. Set month = 0 on the day of first visit
  +  int_mo = months after last visit at the center. 
  +  visitnum = Visit number of the specific subject
  +  datevisit= date of each visit
  +  DOB = date of birth (not shown)
  +  A1C = glycosylated hemoglobin level (marker of diabetes status)
  +  race1 = race of the patient </small>

- **critical feature** of the data:
  +  1. A patient *visits at random times.*
  +  2. A patient *visits random number of times.*
  +  3. *Total length of time between visits* is also random.

#### Research Question
Let's assume: **The physician has a patient currently on a check-up visit.** <br>
- The state for the patient now is **pre-diabetes**.  <br>
- What are the chances s/he will have **diabetes**  <br>
  + in a month? or  <br>
  + a year?  <br>
- How this probability is modulated by her/his **race and gender or risk score**.

#### Data Dimensions
```{r 2, warning= FALSE, message=FALSE}
dim(dpm)
```

- Total of 722 observations and 14 variables

### Exploratory Analysis{.tabset}
#### visits by patient
```{r 4, warning= FALSE, message=FALSE}
DPM<-ddply(dpm, "ID_hash", summarise, Visits=length(ID))
visits_summary <- data.frame(
  Statistic = "Visits per patient",
  Value = sprintf("%.2f +/- %.2f", mean(DPM$Visits), sd(DPM$Visits))
)

knitr::kable(
  visits_summary,
  col.names = c("Statistic", "Value"),
  caption = "Summary of visit counts per patient."
)
```

- An average of 3 visits per patient

#### max/worse state by patient
```{r 5, warning= FALSE, message=FALSE}
Finalstate<-ddply(dpm, "ID_hash", summarise, Final=max(status))
Finalstate %>%
  dplyr::mutate(
    Final = dplyr::recode(
      as.character(Final),
      `1` = "Normoglycemic",
      `2` = "Prediabetes",
      `3` = "Diabetes"
    )
  ) %>%
  dplyr::rename(
    hashed_ID = ID_hash
  ) %>%
  head()
```

- **1 = normal** **2 = pre-diabetes** **3 = diabetes**

#### max/worse state counts: 
```{r 5a, warning= FALSE, message=FALSE}
Finalstate<-ddply(dpm, "ID_hash", summarise, Final=max(status))
table(Finalstate$Final)
```

- **12 subjects remained normoglycemic**, **37 subjects became / remained pre-diabetic from normoglycemic / prediabetic states**, **193 subjects had diabetes at one-point at least**

### Empirical Analysis{.tabset}
#### one-step transitions: counts
```{r 6, warning= FALSE, message=FALSE}
library(msm)
Trans1<-statetable.msm(status, ID_hash, data=dpm)
rownames(Trans1) <- c("From Normal", "From Pre-diabetes", "From Diabetes")
colnames(Trans1) <-c("To Normal", "To Pre-diabetes", "To Diabetes")
Trans1
```

- Does not accounts for predictors: Count one-step transitions ...What was the stage at the next visit <br>
- The table shows the one-step transition counts for all the 242 patients in the dataset <br>
- In all there were 484 transitions. The topleft to downright diagonal and middle column is dominant. <br>

#### one-step transitions: Sankey diagram
```{r 6s, warning= FALSE, message=FALSE}
sankey_links <- expand.grid(
  from_label = rownames(Trans1),
  to_label = colnames(Trans1),
  KEEP.OUT.ATTRS = FALSE,
  stringsAsFactors = FALSE
) %>%
  dplyr::mutate(
    weight = as.vector(Trans1),
    from = dplyr::recode(
      from_label,
      `From Normal` = "current_normal",
      `From Pre-diabetes` = "current_prediabetes",
      `From Diabetes` = "current_diabetes"
    ),
    to = dplyr::recode(
      to_label,
      `To Normal` = "next_normal",
      `To Pre-diabetes` = "next_prediabetes",
      `To Diabetes` = "next_diabetes"
    ),
    source_rank = dplyr::case_when(
      from == "current_normal" ~ 1L,
      from == "current_prediabetes" ~ 2L,
      from == "current_diabetes" ~ 3L
    ),
    target_rank = dplyr::case_when(
      to == "next_normal" ~ 1L,
      to == "next_prediabetes" ~ 2L,
      to == "next_diabetes" ~ 3L
    ),
    link_group = dplyr::case_when(
      to == "next_diabetes" ~ "to_diabetes",
      TRUE ~ "other"
    )
  ) %>%
  dplyr::group_by(from) %>%
  dplyr::mutate(
    pct_source = 100 * weight / sum(weight),
    count_label = paste0(weight, " transitions"),
    pct_label = paste0(round(pct_source, 1), "% of source state"),
    label_pos = dplyr::case_when(
      from == "current_normal" & to == "next_normal" ~ 0.88,
      from == "current_prediabetes" & to == "next_normal" ~ 0.88,
      from == "current_normal" & to == "next_prediabetes" ~ 0.90,
      from == "current_diabetes" & to == "next_prediabetes" ~ 0.94,
      target_rank < source_rank ~ 0.35,
      target_rank > source_rank ~ 0.88,
      TRUE ~ 0.88
    ),
    label_dy = dplyr::case_when(
      from == "current_normal" & to == "next_normal" ~ 18,
      from == "current_prediabetes" & to == "next_normal" ~ 26,
      from == "current_normal" & to == "next_prediabetes" ~ -10,
      from == "current_diabetes" & to == "next_prediabetes" ~ 12,
      target_rank == 1L ~ -2,
      target_rank == 2L ~ 10,
      target_rank == 3L ~ 24
    )
  ) %>%
  dplyr::ungroup() %>%
  dplyr::filter(weight > 0) %>%
  dplyr::select(from, to, weight, count_label, pct_label, label_pos, label_dy, link_group)

sankey_nodes <- data.frame(
  id = c(
    "current_normal", "current_prediabetes", "current_diabetes",
    "next_normal", "next_prediabetes", "next_diabetes"
  ),
  name = c(
    "Current: Normal", "Current: Pre-diabetes", "Current: Diabetes",
    "Next: Normal", "Next: Pre-diabetes", "Next: Diabetes"
  ),
  stringsAsFactors = FALSE
)

if (knitr::is_html_output() && requireNamespace("networkD3", quietly = TRUE)) {
  sankey_links_d3 <- sankey_links %>%
    dplyr::mutate(
      source = match(from, sankey_nodes$id) - 1L,
      target = match(to, sankey_nodes$id) - 1L,
      value = weight
    ) %>%
    dplyr::select(source, target, value, count_label, pct_label, link_group)

  sankey_plot <- networkD3::sankeyNetwork(
    Links = sankey_links_d3,
    Nodes = sankey_nodes,
    Source = "source",
    Target = "target",
    Value = "value",
    NodeID = "name",
    units = "patients",
    fontSize = 14,
    nodeWidth = 30,
    sinksRight = TRUE
  )

  htmlwidgets::onRender(
    sankey_plot,
    '
    function(el, x, data) {
      setTimeout(function() {
        var svg = d3.select(el).select("svg");
        var links = svg.selectAll(".link");
        var lookup = {};

        links
          .style("stroke", function(d) {
            return d.target.name === "Next: Diabetes"
              ? "rgba(220, 38, 38, 0.50)"
              : d.target.name === "Next: Normal"
                ? "rgba(34, 197, 94, 0.42)"
                : "rgba(140, 140, 140, 0.50)";
          })
          .style("stroke-opacity", 1);

        data.keys.forEach(function(key, idx) {
          lookup[key] = {
            pct: data.pcts[idx],
            count: data.counts[idx],
            pos: data.pos[idx],
            dy: data.dy[idx]
          };
        });

        links.each(function(d, i) {
          var key = d.source.name + "|" + d.target.name;
          var meta = lookup[key] || { pct: "", count: d.value };

          d3.select(this).select("title").remove();
          d3.select(this)
            .append("title")
            .text(
              d.source.name + " -> " + d.target.name +
              "\\nCount: " + meta.count +
              "\\nShare of source: " + meta.pct
            );
        });

        svg.selectAll(".link-label-layer").remove();

        var labelLayer = svg.append("g")
          .attr("class", "link-label-layer");

        labelLayer.selectAll("text")
          .data(links.data())
          .enter()
          .append("text")
          .attr("x", function(d, i) {
            var key = d.source.name + "|" + d.target.name;
            var meta = lookup[key] || { pos: 0.5, dy: 0 };
            var path = links.nodes()[i];
            var point = path.getPointAtLength(path.getTotalLength() * meta.pos);
            return point.x;
          })
          .attr("y", function(d, i) {
            var key = d.source.name + "|" + d.target.name;
            var meta = lookup[key] || { pos: 0.5, dy: 0 };
            var path = links.nodes()[i];
            var point = path.getPointAtLength(path.getTotalLength() * meta.pos);
            return point.y + meta.dy;
          })
          .attr("text-anchor", "middle")
          .attr("dy", "0.1em")
          .style("font-size", "11px")
          .style("font-weight", "600")
          .style("fill", "#1f2937")
          .style("paint-order", "stroke")
          .style("stroke", "white")
          .style("stroke-width", "3px")
          .style("stroke-linejoin", "round")
          .style("pointer-events", "none")
          .text(function(d) {
            var key = d.source.name + "|" + d.target.name;
            return (lookup[key] || {}).pct || "";
          });
      }, 300);
    }
    ',
    data = list(
      keys = paste(
        sankey_nodes$name[match(sankey_links$from, sankey_nodes$id)],
        sankey_nodes$name[match(sankey_links$to, sankey_nodes$id)],
        sep = "|"
      ),
      pcts = sankey_links$pct_label,
      counts = sankey_links$weight,
      pos = sankey_links$label_pos,
      dy = sankey_links$label_dy
    )
  )
} else {
  knitr::kable(
    sankey_links %>% dplyr::select(from, to, weight, pct_label),
    col.names = c("From", "To", "Observed transitions", "Percent of source state"),
    caption = "Observed one-step transitions used in the Sankey diagram. Install the networkD3 package and render to HTML to view the interactive Sankey plot."
  )
}
```

- The Sankey diagram visualizes observed patient flow from the current state to the next recorded state.
- It is based on empirical one-step transition counts and does not account for predictors or time between visits.

#### one-step transitions: percentages
```{r 6a, warning= FALSE, message=FALSE}
Trans2<-100*round(prop.table(Trans1,1),3)
Trans2
```
- If the patient is in **Normoglycemic** State currently, the chances that s/he will remain **Normoglycemia** are *62%*.
- If the patient is in **Pre-diabetic** State currently, the chances that s/he will remain **Pre-diabetic state** are *60%*.
- If the patient is in **Diabetic** State currently, the chances that s/he will remain ""Diabetec** are *80.5%*.
- **Note** Empirical analysis does not take into account **predictors** and **time**.

### Model Building{.tabset}
#### Fit a Markov process model {.tabset}
##### Initial Generator for Iteration
```{r 7, warning= FALSE, message=FALSE}
Q <- rbind(c(0, 0.25, 0.5), c(0.166, 0, 0.5), c(0.16, 0.25, 0) )
Q
```
*Initializing the Q matrix. The diagonals set at zero. R-program figures out the actual values.*

##### Specifying initial values 
```{r 8, warning= FALSE, message=FALSE}
Q.crude<-crudeinits.msm(status~cumint, ID_hash, data = dpm, qmatrix = Q)
Q.crude
```

- to start searching for maximum; no covariates are entertained at this time.
- The analysis is purely based on data with state and duration (months) since last visit. The matrix provides rates of decay needed for estimating transition probability matrices.*

##### Results: Estimates
```{r 9a, warning= FALSE, message=FALSE}
dpm.msm <- msm(status ~ cumint, subject = ID_hash, data= dpm, qmatrix = Q, deathexact = FALSE)
dpm.msm
```

- *From the estimated intensities, we see that prediabetic patients (State 2) are 18 times as likely to progress to diabetes (transitions from State 2 to State 3, vs from State 2 to State 1).*

#### Model
```{r 9c, warning= FALSE, message=FALSE}
# pmatrix.msm(dpm.msm, t=.1)
pmatrix.msm(dpm.msm, t=1.0)
```

- **Producing the maximum likelihood estimate of the infinitesimal generator Q.**
  + *Estimate one-step transition probability matrix for any given time. What is the conditional probability of becoming diabetic in one year's time given that the patient is currently in prediabetic state: 0.59*

#### Add Covariates{.tabset} 
##### Covariate: Risk Score **RAF_RS**
```{r 9d, warning= FALSE, message=FALSE}
tidy_qmatrix <- function(qmatrix_obj) {
  q_estimate <- if (is.list(qmatrix_obj) && !is.null(qmatrix_obj$estimate)) {
    qmatrix_obj$estimate
  } else {
    qmatrix_obj
  }

  as.data.frame(as.table(as.matrix(q_estimate)), responseName = "intensity") %>%
    dplyr::rename(from_state_raw = Var1, to_state_raw = Var2) %>%
    dplyr::mutate(
      from_id = as.integer(from_state_raw),
      to_id = as.integer(to_state_raw)
    ) %>%
    dplyr::filter(from_id != to_id) %>%
    dplyr::mutate(
      from_state = dplyr::recode(
        as.character(from_state_raw),
        `1` = "Normoglycemia",
        `State 1` = "Normoglycemia",
        `2` = "Pre-diabetes",
        `State 2` = "Pre-diabetes",
        `3` = "Diabetes",
        `State 3` = "Diabetes",
        .default = as.character(from_state_raw)
      ),
      to_state = dplyr::recode(
        as.character(to_state_raw),
        `1` = "Normoglycemia",
        `State 1` = "Normoglycemia",
        `2` = "Pre-diabetes",
        `State 2` = "Pre-diabetes",
        `3` = "Diabetes",
        `State 3` = "Diabetes",
        .default = as.character(to_state_raw)
      ),
      transition = paste(from_state, "->", to_state)
    ) %>%
    dplyr::select(from_id, to_id, transition, intensity)
}

dpmraf.msm<-msm(status~cumint, subject=ID_hash, data=dpm, qmatrix=Q, deathexact = FALSE, covariates=~RAF_RS)

low_risk_q <- qmatrix.msm(dpmraf.msm, covariates = list(RAF_RS = 0.3), ci = "none")
high_risk_q <- qmatrix.msm(dpmraf.msm, covariates = list(RAF_RS = 1.5), ci = "none")

raf_effect_tbl <- tidy_qmatrix(low_risk_q) %>%
  dplyr::rename(`RAF_RS = 0.3` = intensity) %>%
  dplyr::left_join(
    tidy_qmatrix(high_risk_q) %>%
      dplyr::rename(`RAF_RS = 1.5` = intensity),
    by = c("from_id", "to_id", "transition")
  ) %>%
  dplyr::mutate(
    `1.5 / 0.3 rate ratio` = `RAF_RS = 1.5` / `RAF_RS = 0.3`
  ) %>%
  dplyr::arrange(from_id, to_id) %>%
  dplyr::select(transition, `RAF_RS = 0.3`, `RAF_RS = 1.5`, `1.5 / 0.3 rate ratio`)

gt::gt(raf_effect_tbl) %>%
  gt::tab_header(
    title = "Effect of Risk Score on Transition Intensities",
    subtitle = "RAF_RS = 0.3 versus RAF_RS = 1.5"
  ) %>%
  gt::cols_label(
    transition = "Transition",
    `RAF_RS = 0.3` = "RAF_RS = 0.3",
    `RAF_RS = 1.5` = "RAF_RS = 1.5",
    `1.5 / 0.3 rate ratio` = "1.5 / 0.3 HR"
  ) %>%
  gt::tab_spanner(
    label = "Estimated intensity",
    columns = c(`RAF_RS = 0.3`, `RAF_RS = 1.5`)
  ) %>%
  gt::fmt_number(columns = c(`RAF_RS = 0.3`, `RAF_RS = 1.5`), decimals = 3) %>%
  gt::fmt_number(columns = `1.5 / 0.3 rate ratio`, decimals = 2) %>%
  gt::cols_align(
    align = "center",
    columns = c(`RAF_RS = 0.3`, `RAF_RS = 1.5`, `1.5 / 0.3 rate ratio`)
  ) %>%
  gt::opt_row_striping() %>%
  gt::tab_source_note(
    source_note = "Off-diagonal entries from qmatrix.msm(); larger values indicate faster transitions between states."
  )
```

- *The table summarizes the off-diagonal transition intensities at two illustrative RAF_RS values. A 1.5 / 0.3 HR greater than 1 indicates faster transition rates at the higher risk score.*

##### Model with constrained covariate effects{.tabset}
###### Effect of Race (baseline=AA;)
```{r 9e, warning= FALSE, message=FALSE}
#dpmrace.msm<-msm(status~cumint, subject=ID_hash, data=dpm, qmatrix=Q, deathexact = FALSE, covariates=list("2-3"=~race1, "2-1"=~race1))
dpmrace.msm<-msm(status~cumint, subject=ID_hash, data=dpm, qmatrix=Q, deathexact = FALSE, covariates=~race1, constraint = list(race1=c(1,2,3,1,2,3)))

aa_q <- qmatrix.msm(dpmrace.msm, covariates = list(race1=1), ci = "none")
non_aa_q <- qmatrix.msm(dpmrace.msm, covariates = list(race1=0), ci = "none")

race_effect_tbl <- tidy_qmatrix(aa_q) %>%
  dplyr::rename(AA = intensity) %>%
  dplyr::left_join(
    tidy_qmatrix(non_aa_q) %>%
      dplyr::rename(`Non-AA` = intensity),
    by = c("from_id", "to_id", "transition")
  ) %>%
  dplyr::mutate(
    `Non-AA/AA rate ratio` = `Non-AA` / AA
  ) %>%
  dplyr::arrange(from_id, to_id) %>%
  dplyr::select(transition, AA, `Non-AA`, `Non-AA/AA rate ratio`)

gt::gt(race_effect_tbl) %>%
  gt::tab_header(
    title = "Effect of Race on Transition Intensities",
    subtitle = "AA (race1 = 1) versus non-AA (race1 = 0)"
  ) %>%
  gt::cols_label(
    transition = "Transition",
    AA = "AA",
    `Non-AA` = "Non-AA",
    `Non-AA/AA rate ratio` = "Non-AA/AA HR"
  ) %>%
  gt::tab_spanner(
    label = "Estimated intensity",
    columns = c(AA, `Non-AA`)
  ) %>%
  gt::fmt_number(columns = c(AA, `Non-AA`), decimals = 3) %>%
  gt::fmt_number(columns = `Non-AA/AA rate ratio`, decimals = 2) %>%
  gt::cols_align(align = "center", columns = c(AA, `Non-AA`, `Non-AA/AA rate ratio`)) %>%
  gt::opt_row_striping() %>%
  gt::tab_source_note(
    source_note = "Off-diagonal entries from qmatrix.msm(); larger values indicate faster transitions between states."
  )
```
*The table summarizes the off-diagonal transition intensities by race. A Non-AA/AA HR greater than 1 indicates faster transition rates in the non-AA group.*

###### Effect of Gender (F vs. M)
```{r 9f, warning= FALSE, message=FALSE}
dpmgender.msm<-msm(status~cumint, subject=ID_hash, data=dpm, qmatrix=Q, deathexact = FALSE, covariates=~gender1, constraint = list(gender1=c(1,2,3,1,2,3)))

female_q <- qmatrix.msm(dpmgender.msm, covariates = list(gender1=0), ci = "none")
male_q <- qmatrix.msm(dpmgender.msm, covariates = list(gender1=1), ci = "none")

gender_effect_tbl <- tidy_qmatrix(female_q) %>%
  dplyr::rename(Female = intensity) %>%
  dplyr::left_join(
    tidy_qmatrix(male_q) %>%
      dplyr::rename(Male = intensity),
    by = c("from_id", "to_id", "transition")
  ) %>%
  dplyr::mutate(
    `Male/Female rate ratio` = Male / Female
  ) %>%
  dplyr::arrange(from_id, to_id) %>%
  dplyr::select(transition, Female, Male, `Male/Female rate ratio`)

gt::gt(gender_effect_tbl) %>%
  gt::tab_header(
    title = "Effect of Gender on Transition Intensities",
    subtitle = "Female (gender1 = 0) versus Male (gender1 = 1)"
  ) %>%
  gt::cols_label(
    transition = "Transition",
    Female = "Female",
    Male = "Male",
    `Male/Female rate ratio` = "Male/Female HR"
  ) %>%
  gt::tab_spanner(
    label = "Estimated intensity",
    columns = c(Female, Male)
  ) %>%
  gt::fmt_number(columns = c(Female, Male), decimals = 3) %>%
  gt::fmt_number(columns = `Male/Female rate ratio`, decimals = 2) %>%
  gt::cols_align(align = "center", columns = c(Female, Male, `Male/Female rate ratio`)) %>%
  gt::opt_row_striping() %>%
  gt::tab_source_note(
    source_note = "Off-diagonal entries from qmatrix.msm(); larger values indicate faster transitions between states."
  )
```
*The table summarizes the off-diagonal transition intensities by sex. A Male/Female HR greater than 1 indicates faster transition rates among males.*

###### Effect of Gender (F vs. M) & Race (baseline=AA)
```{r 9g, warning= FALSE, message=FALSE}
dpmgenderrace.msm <-msm(status~cumint, subject=ID_hash, data=dpm, qmatrix=Q, deathexact = FALSE, 
                        covariates=~gender1 + race1, constraint = list(gender1=c(1,2,3,1,2,3)))

female_aa_q <- qmatrix.msm(dpmgenderrace.msm, covariates = list(gender1=0, race1=1), ci = "none")
male_non_aa_q <- qmatrix.msm(dpmgenderrace.msm, covariates = list(gender1=1, race1=0), ci = "none")

genderrace_effect_tbl <- tidy_qmatrix(female_aa_q) %>%
  dplyr::rename(`Female, AA` = intensity) %>%
  dplyr::left_join(
    tidy_qmatrix(male_non_aa_q) %>%
      dplyr::rename(`Male, non-AA` = intensity),
    by = c("from_id", "to_id", "transition")
  ) %>%
  dplyr::mutate(
    `Male non-AA/Female AA rate ratio` = `Male, non-AA` / `Female, AA`
  ) %>%
  dplyr::arrange(from_id, to_id) %>%
  dplyr::select(transition, `Female, AA`, `Male, non-AA`, `Male non-AA/Female AA rate ratio`)

gt::gt(genderrace_effect_tbl) %>%
  gt::tab_header(
    title = "Effect of Gender and Race on Transition Intensities",
    subtitle = "Female, AA (gender1 = 0, race1 = 1) versus Male, non-AA (gender1 = 1, race1 = 0)"
  ) %>%
  gt::cols_label(
    transition = "Transition",
    `Female, AA` = "Female, AA",
    `Male, non-AA` = "Male, non-AA",
    `Male non-AA/Female AA rate ratio` = "Male non-AA/Female AA HR"
  ) %>%
  gt::tab_spanner(
    label = "Estimated intensity",
    columns = c(`Female, AA`, `Male, non-AA`)
  ) %>%
  gt::fmt_number(columns = c(`Female, AA`, `Male, non-AA`), decimals = 3) %>%
  gt::fmt_number(columns = `Male non-AA/Female AA rate ratio`, decimals = 2) %>%
  gt::cols_align(
    align = "center",
    columns = c(`Female, AA`, `Male, non-AA`, `Male non-AA/Female AA rate ratio`)
  ) %>%
  gt::opt_row_striping() %>%
  gt::tab_source_note(
    source_note = "Off-diagonal entries from qmatrix.msm(); larger values indicate faster transitions between states."
  )
```
*The table compares the off-diagonal transition intensities for the two covariate profiles used in the model call. A ratio greater than 1 indicates faster transition rates in the Male, non-AA profile.*
